---
title: "README"
author: "Leonardo R Alves"
date: "November 19, 2015"
output: html_document
---
# Introduction

This course project assignment requires write an R Script. The objective is to demonstrate ability to collect, work with, and clean a data set (Human Activity Recognition Using Smartphones Data Set, more information in the CodeBook.md file) and then, prepare tidy data that can be used for later analysis. To achieve this objective, I just develop an R script (run_analysis.R). See in the next topics the step-by-step process of the R script generated.

##The R script does the following:
* 1 - Merges the training and the test sets to create one data set.
* 2 - Extracts only the measurements on the mean and standard deviation for each measurement.
* 3 - Uses descriptive activity names to name the activities in the data set
* 4 - Appropriately labels the data set with descriptive variable names. 
* 5 - From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

Most of the learnig used to develop this project was found in this article:
https://thoughtfulbloke.wordpress.com/2015/09/09/getting-and-cleaning-the-assignment/

##Importing project data sets
###Reading and downloading Data from the internet.

First of all, set the Work Directory as you prefer using setwd() command.

Here are the data for the project:

https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

```{r eval=FALSE}
download.file("https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip", "./data.zip")
```

Extracting files and folders of zip file downloaded to the Working Directory.

```{r eval=FALSE}
unzip("data.zip")
```

Reading all important .txt files of extracted file "data.zip".

```{r eval=FALSE}
features <- read.table("features.txt")
activity_labels <- read.table("activity_labels.txt")
subject_test <- read.table("subject_test.txt")
x_test <- read.table("X_test.txt")
y_test <- read.table("y_test.txt")
subject_train <- read.table("subject_train.txt")
x_train <- read.table("X_train.txt")
y_train <- read.table("y_train.txt")
```
initializing some R libraries used in this script
```{r eval=FALSE}
library(dplyr)
library(tidyr)
library(gdata)
```
##1 - Merging training and test sets to create one data set for each.

###Training set
```{r eval=FALSE}
y_train <- rename(y_train, activity = V1)
subject_train <- rename(subject_train, id = V1)
tmp <- bind_cols(y_train, x_train)
training <- bind_cols(subject_train, tmp)
```
###Test set
```{r eval=FALSE}
y_test <- rename(y_test, activity = V1)
subject_test <- rename(subject_test, id = V1)
tmp2 <- bind_cols(y_test, x_test)
test <- bind_cols(subject_test, tmp2)
```
Merging the data sets training and test in the total data set

```{r eval=FALSE}
total <- bind_rows(training, test)
```

##2 - Extracts only the measurements on the mean and standard deviation for each measurement. 

Puting the labels in the variables of "total" dataset using "features" dataset. Indentifying the indexes of duplicated (if exist) labels in "features".
Vector dup receiving duplicated labels. If duplicated, then number 1 is assigned, else 0.
```{r eval=FALSE}
dup <- as.numeric(duplicated(features$V2))
ind <-vector() #initializing vector ind
j <- 0
for (i in 1:561){ #number of labels in the features$V2
  if (dup[i]==1){ # if equal 1, means that there is a duplicated label
     j <- j+1
     ind[j] <- i # vector that contains the indexes of duplicated labels in features.
  }
}
```
There is some duplicated labels. Removing duplicated labels from features using unique function
```{r eval=FALSE}
features_nodup <- unique(features$V2)
```
Using the indexes of duplicated features of ind vector to exclude them from dataframe total.
```{r eval=FALSE}
total_nodup <- total[,-c(319:346,398:425, 477:504)]
# total_nodup: total with no duplicated label
```
Applying the correct labels to the variables in total_nodup dataset.
```{r eval=FALSE}
for (i in 3:479){ # start in 3 because the first two factor are named already
  names(total_nodup)[i] <- as.character(features_nodup[i-2,1])
}
```
Checking new labels of variables.
```{r eval=FALSE}
names(total_nodup)
```
Selecting only variables related to the "mean" and "std".
```{r eval=FALSE}
tmp_mean <- select(total_nodup, c(id, activity, contains("mean", ignore.case = TRUE)))
tmp_std <- select(total_nodup, c(id, activity, contains("std", ignore.case = TRUE)))
```
Bind "mean" and "std" variables. The "id" and "activity" variables of tmp_std dataset was excluded (duplicated). The total_final dataset was created.
```{r eval=FALSE}
total_final <- bind_cols(tmp_mean, tmp_std[,-c(1,2)])
```

## 3 - Uses descriptive activity names to name the activities in the data set.

Using recode function of library(car) to recode the numbers of activity using the names in activity_label dataset.
```{r eval=FALSE}
library(car)
total_final$activity <- recode(total_final$activity, "1 = 'WALKING'; 2 = 'WALKING_UPSTAIRS'; 3 = 'WALKING_DOWNSTAIRS'; 4 = 'SITTING'; 5 = 'STANDING'; 6 = 'LAYING'")
```

## 4 - Appropriately labels the data set with descriptive variable names.

Storing the labels of total_final dataset in a new df actual_names.
```{r eval=FALSE}
actual_names <- as.data.frame(names(total_final))
```
Export actual_names to .csv file. (reason: This will help to create a new file with descriptive variable names).
```{r eval=FALSE}
write.csv(actual_names, "actual_names.csv")
```
Manualy edit two new files in libreoffice calc, then save in .csv files.

First, the codification table (table to relate the old labels with the new ones) read this file and store in cod_table dataset.
```{r eval=FALSE}
cod_table <- read.csv(codification_of_labels_table.csv, header = TRUE)
```
Second, separate only the columns with the new labels read this file and store in new_names dataset.
```{r eval=FALSE}
new_names <- read.csv("new_names.csv", header = TRUE)
```
Rename the total_final dataset using new_names.
```{r eval=FALSE}
for (i in 1:88){ # 88 is the total of variables of total_final dataset
  names(total_final)[i] <- as.character(new_names[i,1])
}
```
Determining the factors in the total_final dataset.
```{r eval=FALSE}
total_final$id_subject <- as.factor(total_final$id_subject)
total_final$type_of_activity <- as.factor(total_final$type_of_activity)
```
Finally the tidy dataset of step 4 is ready.
```{r eval=FALSE}
view(total_final)
```

##5 - From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

Using function ddply of library "plyr" to help in this part.
```{r eval=FALSE}
library(plyr)

# The tidy dataset with the average of each variable for each activity and each subject.

HAR_Smartphones_tidy <- ddply(total_final, .(id_subject, type_of_activity), function(x) colMeans(x[, - (1:2)]))
```
Writing the tidy dataset to a .txt file "HAR_Smartphones_tidy.txt"
```{r eval=FALSE}
write.table(HAR_Smartphones_tidy, file = "HAR_Smartphones_tidy.txt", row.name = FALSE)
```
The end.